Icicle: write once, run once

### 2016-06-07 2016fhpc/section/02-Source.tex
Wibbles on S3.1

The grammar for Icicle is given in Figure~\ref{fig:source:grammar}.
Value types $T$ include numbers, booleans or maps.
Modal types $\TauMode$ include the pure value types, and modalities associated with a value type.
Function types $\TauFun$ include non-function modal types, or a function from modal type to function type.
As Icicle is a first-order language, function types are not value types.

Table definitions $\mi{Table}$ define a table name and the names and types of columns.
Expressions $\mi{Exp}$ can be variable names, constants, applications of primitives or functions.
The @fold@ construct defines the name of an accumulator, the expression for the initial value, and the expression used to update the accumulator for each element of the stream.
The @filter@ construct defines a predicate and an expression to accumulate values for which the predicate is true.
The @group@ construct defines an expression used to determine the key for each element of the stream, and an expression to accumulate the values that share a common key.

@@

% ---------------------------------------------------------
\subsection{Type system}
The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
The judgment form $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ associates an expression $\mi{Exp}$ with its type $M$ under context $\Gamma$.
The judgment form $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.

Rules TcNat, TcBool, TcMap and TcPair assign types to literal values.
Rule TcVar performs variable lookup in the context.
Rule TcBox assigns a pure expression either @Element@ or @Aggregate@ type.

% shows that a pure value can be converted to an @Element@ or @Aggregate@, by either replicating it along the stream or returning it at the end of the stream.
% The premise $\tau~\in~T$ requires that $\tau$ is a pure value, rather than an @Element@ or an @Aggregate@.

Rules TcPrimApp and TcFunApp produce the type of the a primitive or function applied to its arguments. Rule TcLet is standard.

% find the result type of primitive and function application.
% They first find the type of the primitive or function, and the types of the arguments, then use the application boxing/unboxing judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ to find the return type and modality of the application.
% The set-comprehension syntax $\{e_i\}$ means that this rule allows any number of arguments.
% Rule TcLet is a standard let rule, finding the type of the bound expression and inserting it into the context when typechecking the remaining expression.

@@

% The initial value of the fold must be pure.
% The update has the current value of the fold inserted in its context as an @Element@, and the update must be an @Element@.
% The final return type of the fold is an @Aggregate@.

% Rule TcFilter requires its predicate to be a stream of booleans ($@Element@~\BB$).
% The rest of the filter must be an @Aggregate@; returning an @Element@ stream with a different rate is not allowed, as explained in\sref{s:IcicleSource:bounded}.
% Rule TcGroup is similar to filter, except returning a @Map@.

The rules PrimArith, PrimRel, PrimLookup, PrimFst and PrimSnd assign types to primitives.
Rule AppArgs produces the type of a function or primitive applied to its arguments.
Rule AppRebox is used when the arguments have modal type $m$ --- applying a pure function to arguments of mode $m$ produces a result of the same mode.

@@

% The next judgment shows how function application boxes and unboxes modalities. This judgment takes a function type and all its arguments, and returns the result type of the function.

% Rule AppArgs applies when a function is applied to exactly the right arguments, and simply returns the result of the function.

% Rule AppRebox is used when a pure function is applied to a non-pure value.
% If the arguments and the result type are all pure, then the function can be applied to arguments of a particular modality, and the return type becomes that modality too.
% Note that rule TcBox can also apply implicit boxing here; when only some of the arguments are non-pure, the pure ones can be boxed.
% These rules are tricky to get right; an earlier version allowed applying a modal argument to a pure argument if the return type fitted that mode, which can lead to giving the results of folds as initial values for folds.

% inserts the argument types into the context and returns a new context with the function's type, which is the type of the arguments to the result type.

% Rule CheckQuery checks a query's types, and makes sure it returns an @Aggregate@.

% It first creates a new context with the table columns bound to @Element@s, and typechecks each function.
% Then, each query is checked, with the table columns and all functions available.
% The final returned context is all query names and their types.

% ---------------------------------------------------------
\subsection{Evaluation}
The evaluation rules for Icicle are given in figure~\ref{fig:source:eval}.
Grammar $N$ defines the modes of evaluation, including pure computation.
Grammar $\Sigma$ defines a heap containing stream values.
Grammar $V'$ defines a different type of stream value for each evaluation mode:
\begin{itemize}
\item
@Pure@ computations are represented as a single value;
\item
@Element@ stream transformers are represented as a function applied to each element; and
\item
@Aggregate@ folds are represented as an initial (zero) state, an update (konstrukt) function applied to each stream element and current state, and an eject function applied to the final state.
\end{itemize}

>>>>

The grammar for Icicle is given in Figure~\ref{fig:source:grammar}.
Value types $T$ include numbers, booleans and maps.
Modal types $\TauMode$ include the pure value types, and modalities associated with a value type.
Function types $\TauFun$ include non-function modal types, and functions from modal type to function type.
As Icicle is a first-order language, function types are not value types.

Table definitions $\mi{Table}$ define a table name and the names and types of columns.
Expressions $\mi{Exp}$ include variable names, constants, applications of primitives and functions.
The @fold@ construct defines the name of an accumulator, the expression for the initial value, and the expression used to update the accumulator for each element of the stream.
The @filter@ construct defines a predicate and an expression to accumulate values for which the predicate is true.
The @group@ construct defines an expression used to determine the key for each element of the stream, and an expression to accumulate the values that share a common key.

@@

% ---------------------------------------------------------
\subsection{Type System}
The typing rules for Icicle are given in Figure~\ref{fig:source:type:exp}.
The judgment form $\Typecheck{\Gamma}{\mi{e}}{\TauMode}$ associates an expression $\mi{e}$ with its type $M$ under context $\Gamma$.
The judgment form $\TypecheckP{\mi{p}}{\TauFun}$ associates a primitive with its function type.
The judgment form $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used to lift function application to modal types: a function type applied to a list of modal argument types produces a result type and matching mode.
The judgment form $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and produces an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and produces a context containing only the types of the queries.

Rules TcNat, TcBool, TcMap and TcPair assign types to literal values.
Rule TcVar performs variable lookup in the context.
Rule TcBox assigns an expression either @Element@ or @Aggregate@ type.

Rules TcPrimApp and TcFunApp produce the type a primitive or function applied to its arguments. Rule TcLet is standard.

@@

Rules PrimArith, PrimRel, PrimLookup, PrimFst and \mbox{PrimSnd} assign types to primitives.
Rule AppArgs produces the type of a function or primitive applied to its arguments.
Rule AppRebox is used when the arguments have modal type $m$ --- applying a function to arguments of mode $m$ produces a result of the same mode.

@@

% ---------------------------------------------------------
\subsection{Evaluation}
Evaluation rules for Icicle are given in Figure~\ref{fig:source:eval}.
Grammar $N$ defines the modes of evaluation, including pure computation.
Grammar $\Sigma$ defines a heap containing stream values.
Grammar $V'$ defines a different type of stream value for each evaluation mode:
\begin{itemize}
\item
@Pure@ computations are represented as a single value;
\item
@Element@ stream transformers are represented as a function applied to each element; and
\item
@Aggregate@ folds are represented as an initial (zero) state, an update (konstrukt) function applied to each stream element and current state, and an eject function applied to the final state.
\end{itemize}
